---
title: "Swimlane gallery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Swimlane gallery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  fig.width = 9,
  fig.height = 5.5,
  out.width = "100%"
)
```

```{r setup}
library(dplyr)
library(ggplot2)
library(ggswimlane)
```

All examples use the bundled `patient_disposition` data, a simulated
disposition table for 20 trial participants.

```{r}
patient_disposition
```

## A basic swimlane

`geom_swimlane()` draws one bar per subject. `order_swimlane()` relevels the
subject column first so lanes run shortest to longest, and
`theme_swimlane()` supplies the finished look. Time axes read best with
breaks at protocol-meaningful intervals, so every example here sets 12-week
breaks rather than accepting the ggplot2 defaults:

```{r basic, fig.alt="Swimlane plot of 20 subjects with blue bars ordered shortest to longest from the bottom"}
patient_disposition |>
  order_swimlane(subject, weeks_on_study) |>
  ggplot() +
  geom_swimlane(subject, weeks_on_study) +
  scale_x_continuous(breaks = scales::breaks_width(12)) +
  labs(x = "Weeks on study") +
  theme_swimlane()
```

Add a fill variable to color lanes by cohort; pass the same variable to
`order_swimlane()` to group the lanes:

```{r cohorts, fig.alt="Swimlane plot with lanes grouped and colored by cohort, blue for Cohort A and orange for Cohort B"}
patient_disposition |>
  order_swimlane(subject, weeks_on_study, cohort) |>
  ggplot() +
  geom_swimlane(subject, weeks_on_study, cohort) +
  scale_x_continuous(breaks = scales::breaks_width(12)) +
  labs(x = "Weeks on study") +
  theme_swimlane()
```

## Annotating lanes

Three layers mark events along each lane, and each generates its legend
entry automatically:

* `geom_swimlane_status()` marks the end of each lane with the subject's
  status; subjects still on study (`NA` status) get a rightward arrow.
* `geom_swimlane_marker()` marks on-lane events such as responses.
* `geom_swimlane_rug()` flags baseline characteristics in the left margin.

```{r annotated, fig.height=6, fig.alt="Swimlane plot colored by cohort with shaped points marking each subject's end-of-study status, partial responses, prior therapy flags in the left margin, and arrows for subjects still on study"}
patient_disposition |>
  mutate(
    prior_drug = if_else(prior_drug == "Yes", "Prior therapy", NA)
  ) |>
  order_swimlane(subject, weeks_on_study, cohort) |>
  ggplot() +
  geom_swimlane(subject, weeks_on_study, cohort) +
  geom_swimlane_status(subject, weeks_on_study, reason_off_study) +
  geom_swimlane_marker(
    subject, partial_response,
    marker_label = "Partial response"
  ) +
  geom_swimlane_rug(subject, prior_drug) +
  scale_x_continuous(breaks = scales::breaks_width(12)) +
  labs(title = "Time on study by subject", x = "Weeks on study") +
  theme_swimlane()
```

## Two-color bars: stacked segments

With one row per subject and phase, bar segments stack in factor-level
order. Here each lane splits into treatment and follow-up phases; note that
`order_swimlane()` runs before pivoting so lanes stay ordered by total
duration.

Shapes assign in factor-level order, so a plot with different status levels
would reshuffle them: this plot has no partial responses, and "Physician
decision" would silently inherit the diamond that meant "Partial response"
above. When a document has several swimlanes, pin each status to its shape
with a named `values` vector:

```{r stacked, fig.alt="Swimlane plot where each bar is split into a blue treatment segment followed by an orange follow-up segment, with the same status shapes as the previous figure"}
status_shapes <- c(
  "Adverse event" = 21,
  "Completed" = 24,
  "Disease progression" = 22,
  "Partial response" = 23,
  "Physician decision" = 25,
  "Prior therapy" = 1,
  "Withdrawal by subject" = 0
)

patient_disposition |>
  order_swimlane(subject, weeks_on_study, cohort) |>
  mutate(
    Treatment = weeks_on_treatment,
    `Follow-up` = weeks_on_study - weeks_on_treatment
  ) |>
  tidyr::pivot_longer(
    c(Treatment, `Follow-up`),
    names_to = "phase", values_to = "weeks"
  ) |>
  mutate(phase = factor(phase, levels = c("Treatment", "Follow-up"))) |>
  ggplot() +
  geom_swimlane(subject, weeks, phase) +
  geom_swimlane_status(subject, weeks_on_study, reason_off_study) +
  scale_shape_swimlane(
    values = status_shapes,
    guide = guide_legend(order = 3, nrow = 2)
  ) +
  scale_x_continuous(breaks = scales::breaks_width(12)) +
  labs(title = "Treatment and follow-up phases", x = "Weeks on study") +
  theme_swimlane()
```

## Text annotations

Prefer text over symbols? `geom_swimlane_label()` writes each status past
the end of its bar. Subjects with `NA` are skipped, which would leave
ongoing subjects indistinguishable from missing data, so recode them to a
label first. Widen the right margin to make room:

```{r labels, fig.alt="Swimlane plot with each subject's status, including On study, written as text just past the end of its bar"}
patient_disposition |>
  mutate(reason_off_study = coalesce(reason_off_study, "On study")) |>
  order_swimlane(subject, weeks_on_study, cohort) |>
  ggplot() +
  geom_swimlane(subject, weeks_on_study, cohort) +
  geom_swimlane_label(subject, weeks_on_study, reason_off_study) +
  scale_x_continuous(breaks = scales::breaks_width(12)) +
  labs(x = "Weeks on study") +
  theme_swimlane(extra_margin_r = 60)
```

## Customization

The built-in scales step aside for any discrete scale you add (ggplot2
prints a message when a scale is replaced). Theme settings pass through
`theme_swimlane()`, including an inside legend:

```{r custom, fig.alt="Swimlane plot with a green and orange Dark2 palette and the cohort legend placed inside the panel at the lower right"}
patient_disposition |>
  order_swimlane(subject, weeks_on_study, cohort) |>
  ggplot() +
  geom_swimlane(subject, weeks_on_study, cohort) +
  scale_fill_brewer(palette = "Dark2") +
  scale_x_continuous(breaks = scales::breaks_width(12)) +
  labs(x = "Weeks on study") +
  theme_swimlane(
    base_size = 12,
    legend.position = "inside",
    legend.position.inside = c(0.9, 0.15)
  )
```
